A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-03-07 papers/contact/paper.tex
a few small changes to paper Tested everything on crystal.
various improvements to contact paper Tested everything on bingley.

\newcommand{\red}[1]{{\bf \color{red} #1}}
\newcommand{\blue}[1]{{\bf \color{blue} #1}}
\newcommand{\green}[1]{{\bf \color{green} #1}}
\newcommand{\rr}{\textbf{r}}
\newcommand{\xx}{\textbf{x}}
\newcommand{\refnote}{\red{[ref]}}

@@

\newcommand{\derivation}[1]{#1} % Use this to show all derivations in detail
%\newcommand{\derivation}[1]{} % Use this for nice pegagogical paper...

@@

There has been considerable recent interest in using Statistical
Associating Fluid Theory (SAFT) to construct classical density
functionals to describe associating
fluids\cite{\saftlocaldft,\saftnonlocaldft}.  This approach has been
successful in qualitatively describing the dependence of surface
tension on temperature.  Unfortunately, most of these constructed
functionals\cite{\saftlocaldft} cannot be used to study arbitrary
inhomogeneous density distributions, because they use local density
approximations that are only valid for densities that vary slowly over
molecular length scales.  A key input to those SAFT-based functionals
that can handle sharply varying densities is the correlation function
evaluated at contact. \fixme{I need to figure out how best to phrase
  this... do we want to compute correlation at contact, or contact
  density itself? I prefer the latter, but the former is what is
  commonly included in formulas.}

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  The chain and association contributions to the SAFT
free energy are of particular interest for this paper, since each uses
as input the contact value of the correlation function.
%
Yu and Wu introduced in 2002 a functional for the association term of
the free energy, which included a functional for the contact value of
the correlation function (see
Section~\ref{sec:yuwu})\cite{yu2002fmt-dft-inhomogeneous-associating},
which has subsequently been used in other
functionals\cite{fu2005vapor-liquid-dft, bryk2006density}.
%% Each of these terms has some dependence on the
%% inhomogeneity in the density.  The ideal gas is purely local, and in
%% fact must be the only purely local term in the free energy in order
%% for the functional to satisfy the contact-value theorem.  The
%% hard-sphere contribution contribution to the free energy has been
%% thoroughly studied, and is well approximated by the White Bear version
%% of the Fundamental-Measure Theory (FMT)
%% functional\cite{roth2002whitebear}.
Two functionals for the chain contribution have recently been
introduced\cite{bryk2006density, gross2009density}, one which uses the
contact value of the correlation function of Yu and
Wu\cite{bryk2006density}, while the other introduces a new functional
for the contact value of the correlation
function (see Section~\ref{sec:gross})\cite{gross2009density}.
%% The dispersion term is long-range and thus has significant
%% dependence on the density distribution, but because its relatively
%% weak position dependence, is amenable to mean-field approximations.
Given these different approaches, it seems worthwhile to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.
\fixme{

Jeff's Notes:
A novice (me) is confused by the use of 'contact value of correlation function' throuhout this introduction while throughout the rest of the paper we refer to the 'contact density'.  Pehaps the audince will understand that they're interchangable (if they are) but this was confusing to me.  Also, I would think I'd like to see something like:

We first discuss two possibilities for the definition of contact density.  The first involves a more symmetric interpretation and so we call it the 'S case'.  The second involves a sphere-centered, assymetrical interpretation and so we call it the 'A case'.  We present where in the SAFT free energy formulation these contact densities will contribute.  We next present a thermodynamic method that we will use to calculate the above contact densities.  This method involves the use of surface pressure and the minimization of the excess free energy refered to above.  We use the White Bear and White Bear Mark II formulations of these free energy terms and take neccesary derivitives in the concpetual spirit of the seperately defined S an A case contact densities.  Finally we compare the resulting contact densities with those of Gross, those of Yu and Wu, and as well with our wn monte carlo simulations.}

\fixme{Should we include these equations? Maybe just later?}
\begin{align}
  \frac{A_\textit{chain}}{kT} &= -(m-1) n \left(\ln\left(g_\sigma\right)-1\right)
\end{align}
\begin{align}
  \frac{A_\textit{assoc}}{kT} &= \sum_i n \left(\ln X_i - \frac12 X_i + \frac12\right) \\
  X_i &= \frac{1}{1 + \sum_j n g_\sigma X_j\kappa_{ij} \left(e^{\beta \epsilon_{ij}}-1\right)}
\end{align}

\section{Contact density and correlation function}

\fixme{We should decide on good names for these two densities.  In
  this section I define them formally, and we should keep using the
  same names (and mathematical symbols) for them throughout.}

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume of
finding one particle at position $\rr_1$ and the other at position
$\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
the diameter $\sigma$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
In a system with an inhomogeneous density, we seek a \emph{local}
value for $g_\sigma$.  There are two reasonable options for defining
such a local function: a symmetric formulation (which we refer to as $S$) and an
asymmetric formulation (which we refer to as $A$). \fixme{Should we
  introduce a third one, for Yu and Wu?}

For the symmetric $S$ case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_S(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \\
  n_S(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_S$ is the same as the density $n_0$
defined in Fundamental Measure Theory (FMT), and seems ideal for
treating touching spheres.

@@

In the process of defining these two averaged correlation functions,
we also defined two averaged densities, $n_S(\rr)$ and $n_A(\rr)$,
which are the density of spheres available to be in
contact.  In the homogeneous limit, $n_S = n_A = n$.  It is an
open question, which of these averages will be more useful in any
particular functional.  Suffice to say, either average is a
\emph{possible} way to convert a function that is defined for a
homogeneous system to a functional that is applicable to inhomogeneous
systems.

Yu and Wu introduce an approximation for
$g^S$\cite{yu2002fmt-dft-inhomogeneous-associating}, which will be
discussed in Section~\ref{sec:yuwu}.  Gross introduced an
approximation for $g^A$\cite{gross2009density}, which will be
introduced in Section~\ref{sec:gross}.  We will also introduce our own
approximation to $g^A$ which is based on the assumption of
thermodynamic consistency within FMT, in
Section~\ref{contact-at-sphere}.  We will report on the accuracy of
these approximations by comparing with Monte Carlo simulations of the
inhomogeneous hard-sphere fluid.  We will focus on the contact
densities, rather than the correlation function itself, because these
are simpler, and because they are the quantity that you want to
``average''.  i.e. if you were to average $g$ itself, you'd want to
weight that average based on the density, which would mean averaging
$n^{(2)}$.

\subsection{Association free energy of SAFT}
\newcommand\epsilonassoc{\ensuremath{\varepsilon_\textit{AB}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{AB}}}
\newcommand\ncontact{\ensuremath{n_\textit{contact}}}

@@

In order to motivate the derivation of our functional for contact
density, we begin by deriving the well-known contact density for a
homogeneous hard-sphere fluid from the Carnahan-Starling free energy.
The contact density is most easily computed from the contact density
theorem, which states that the pressure on any hard surface is
determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact}.
\end{align}
Since we are interested in the contact density at the surface of the
hard spheres, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the Carnahan-Starling free energy on
hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{3} R^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi (2R)^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

We use the White Bear version of the Fundamental-Measure Theory~(FMT)
functional published in reference~\cite{roth2002whitebear}.  The FMT
functional describes the excess free energy of a hard-sphere fluid.
This particular FMT reduces to the Carnahan-Starling equation of state
for homogeneous systems.
\begin{equation}
A_\textit{HS}[n] = k_B T \int \left(\Phi_1(\xx) + \Phi_2(\xx) + \Phi_3(\xx)\right) d\xx \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
  n_3 + (1-n_3)^2 \ln(1-n_3)
}{
  36\pi n_3^2\left( 1 - n_3 \right)^2
} ,
\end{align}
using the weighted densities
\begin{align}
  n_3(\xx) &= \int n(\xx') \Theta(\left|\xx - \xx'\right| - R) d\xx' \\
  n_2(\xx) &= \int n(\xx') \delta(\left|\xx - \xx'\right| - R) d\xx'
\end{align}
\begin{align}
  \mathbf{n}_{V2} &= \mathbf{\nabla} n_3 , \quad
  \mathbf{n}_{V1} = \frac{\mathbf{n}_{V2}}{4\pi R} \\
  n_1 &= \frac{n_2}{4\pi R} , \quad
  n_0 = \frac{n_2}{4\pi R^2}
\end{align}

@@

%% \textcolor{red}{FIXME WRITE THIS SECTION!} The direct correlation
%% function is readily accessible from density-functional theory, as the
%% second functional derivative of the free energy:
%% \textcolor{red}{(Something like...)}
%% \begin{align}
%%   -c^{(2)}(\xx_1, \xx_2) &= \frac{1}{k_BT}\frac{\delta^2 F_{ex}}{\delta n(\xx_1) \delta n(\xx_2)}
%% \end{align}
%% The direct correlation function is related to the ordinary correlation
%% function by the Ornstein-Zernike relation
%% \begin{align}
%%   h(\xx_1, \xx'_2) &= g(\xx_1, \xx_2) - 1 \\
%%   h(\xx_1, \xx_2) &= c(\xx_1, \xx_2)
%%     + \int n(\xx_3)c(\xx_1, \xx_3) h(\xx_2, \xx_3)d\xx''
%% \end{align}
%% From these two equation, we could in principle extract the correlation
%% function evaluated at contact from the DFT free energy functional.
%% The Ornstein-Zernike equation is usually solved by taking a Fourier
%% transform of all the relevant quantities.  In the inhomogeneous case,
%% however, this approach isn't workable, since the Fourier transforms
%% don't speed things up like you'd hope, and we don't have a periodic
%% system in general.

@@

%% One reasonably simple question to ask is what the \emph{mean} contact
%% density of an inhomogeneous system is.  This comes down to simply
%% reproducing the derivation for a homogeneous system using the
%% FMT functional.
%% \begin{align}
%%   p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
%%   \ncontact &= \frac{1}{N k_BT 4\pi (2R)^2} \frac{dA_{HS}}{dR}
%% \end{align}
%% While this should give the true mean contact density for any given
%% system, this is not particularly useful, as we have no way of knowing
%% the contact density on any given sphere.  However, the mean contact
%% density comes out as an integral, so we can plot the integrand, and
%% treat it as a somewhat ill-defined local contact density.

@@

\section{Asymmetrically averaged correlation function}\label{contact-at-sphere}

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{4\pi (2R)^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  \ncontact(\xx) &= \frac{1}{k_BT 4\pi (2R)^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})}
\end{align}
After considerable math to do the above functional derivative, we find
a functional giving the contact density for spheres at any given
position.  This functional unavoidably involves more convolutions than
are present in the free energy itself, which makes it relatively
expensive.

@@

\section{Symmetrically averaged correlation function}

An alternate definition of the averaged correlation function for an
inhomogeneous system would be the correlation function averaged for
spheres \emph{touching at a given point}.  This would be a symmetric
measure, since both spheres involved are treated in the same way.  In
fact, the density of such spheres is just the density $n_0(\xx)$ from
FMT.  See Figure~\ref{fig:n0} for an illustration $n_0$.  Since $n_0$
is the fundamental measure of sphere surface contact, it makes sense
to define a corresponding correlation function $g_\sigma^S$.  This is
\emph{almost} what is done by Yu and
Wu\cite{yu2002fmt-dft-inhomogeneous-associating}, as discussed in
Section~\ref{sec:yuwu}, although they use a modified density $\zeta
n_0$ to account for density gradients.

@@

We performed a Monte-Carlo simulation of the hard sphere fluid to
measure the contact density as a function of position for a few simple
geometries.  For each scenario, we compute and compare several
quantities.  We compare the Monte Carlo density with the DFT
prediction.  We also plot Monte Carlo results for the two contact
densities: the contact density evaluated at contact, and the contact
density evaluated at the sphere location.  Alongside these, we plot
the contact density of Yu and
Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}, the contact density
of Gross~\cite{gross2009density}, and the ``contact density at
sphere'' from Section~\ref{contact-at-sphere}.  In principle, the
functional of Yu (FIXME) should match the CenConDensity, while the DFT
at sphere should match the ConDensity...

@@

\begin{widetext}
\section{The White Bear free energy for homogeneous systems}

>>>>

\newcommand{\red}[1]{{\bf \color{red} #1}}
\newcommand{\blue}[1]{{\bf \color{blue} #1}}
\newcommand{\green}[1]{{\bf \color{green} #1}}
\newcommand{\rr}{\textbf{r}}
\newcommand{\refnote}{\red{[ref]}}

@@

%\newcommand{\derivation}[1]{#1} % Use this to show all derivations in detail
\newcommand{\derivation}[1]{} % Use this for nice pegagogical paper...

@@

There has been considerable recent interest in using Statistical
Associating Fluid Theory (SAFT) to construct classical density
functionals to describe associating
fluids\cite{\saftlocaldft,\saftnonlocaldft}.  This approach has been
successful in qualitatively describing the dependence of surface
tension on temperature.  Unfortunately, most of these constructed
functionals\cite{\saftlocaldft} cannot be used to study arbitrary
inhomogeneous density distributions, because they use local density
approximations that are only valid for densities that vary slowly over
molecular length scales.  A key input to those SAFT-based functionals
that can handle sharply varying densities is the correlation function
evaluated at contact.

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  The chain and association contributions to the SAFT
free energy are of particular interest for this paper, since each uses
as input the contact value of the correlation function.
%
Yu and Wu introduced in 2002 a functional for the association term of
the free energy, which included a functional for the contact value of
the correlation function (see
Section~\ref{sec:yuwu})\cite{yu2002fmt-dft-inhomogeneous-associating},
which has subsequently been used in other
functionals\cite{fu2005vapor-liquid-dft, bryk2006density}.
%% Each of these terms has some dependence on the
%% inhomogeneity in the density.  The ideal gas is purely local, and in
%% fact must be the only purely local term in the free energy in order
%% for the functional to satisfy the contact-value theorem.  The
%% hard-sphere contribution contribution to the free energy has been
%% thoroughly studied, and is well approximated by the White Bear version
%% of the Fundamental-Measure Theory (FMT)
%% functional\cite{roth2002whitebear}.
Two functionals for the chain contribution have recently been
introduced\cite{bryk2006density, gross2009density}, one which uses the
contact value of the correlation function of Yu and
Wu\cite{bryk2006density}, while the other introduces a new functional
for the contact value of the correlation
function (see Section~\ref{sec:gross})\cite{gross2009density}.
%% The dispersion term is long-range and thus has significant
%% dependence on the density distribution, but because its relatively
%% weak position dependence, is amenable to mean-field approximations.
Given these different approaches, it seems worthwhile to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.

We first discuss two possibilities for the definition of contact
density.  The first involves a more symmetric interpretation and so we
call it the '$S$ case'.  The second involves a sphere-centered,
assymetrical interpretation and so we call it the '$A$ case'.  We
present where in the SAFT free energy formulation these contact
densities will contribute.  We next present a thermodynamic method
that we will use to calculate the above contact densities.  This
method involves the use of surface pressure and the minimization of
the excess free energy refered to above.  We use the White Bear and
White Bear Mark II formulations of these free energy terms and take
neccesary derivitives in the concpetual spirit of the seperately
defined $S$ an $A$ case contact densities.  Finally we compare the
resulting contact densities with those of Gross, those of Yu and Wu,
and as well with our own Monte Carlo simulations.

\section{Correlation function with inhomogeneity}

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume of
finding one particle at position $\rr_1$ and the other at position
$\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
the diameter $\sigma$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
In a system with an inhomogeneous density, we seek a \emph{local}
value for $g_\sigma$.  There are two reasonable options for defining
such a local function: a symmetric formulation (which we refer to as $S$) and an
asymmetric formulation (which we refer to as $A$).

For the symmetric $S$ case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_0(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \\
  n_0(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_0$ is one of the fundamental measures
of Fundamental Measure Theory (FMT), and seems ideal for treating
touching spheres.

@@

In the process of defining these two averaged correlation functions,
we also defined two averaged densities, $n_0(\rr)$ and $n_A(\rr)$,
which are the density of spheres available to be in
contact.  In the homogeneous limit, $n_0 = n_A = n$.  It is an
open question, which of these averages will be more useful in any
particular functional.  Suffice to say, either average is a
\emph{possible} way to convert a function that is defined for a
homogeneous system to a functional that is applicable to inhomogeneous
systems.

Yu and Wu introduce an approximation for
$g^S$\cite{yu2002fmt-dft-inhomogeneous-associating}, which we will
discus in Section~\ref{sec:yuwu}.  Gross introduced an
approximation for $g^A$\cite{gross2009density}, which we will
introduce in Section~\ref{sec:gross}.  We will also introduce our own
approximations for both $g^A$ and $g^S$ which are based on the
assumption of thermodynamic consistency within FMT, in
Sections~\ref{sec:g-S} and~\ref{sec:g-A}.  We will report on the accuracy of
these approximations by comparing with Monte Carlo simulations of the
inhomogeneous hard-sphere fluid.  We will focus on the contact
densities, rather than the correlation function itself, because these
are simpler, and because they are the quantity that you want to
``average''.  i.e. if you were to average $g$ itself, you'd want to
weight that average based on the density, which would mean averaging
$n^{(2)}$.

\subsection{Association free energy of SAFT}

\fixme{Integrate the following equations into this section...}
\begin{align}
  \frac{A_\textit{chain}}{kT} &= -(m-1) n \left(\ln\left(n g_\sigma \right)-1\right)
\end{align}
\begin{align}
  \frac{A_\textit{assoc}}{kT} &= \sum_i n \left(\ln X_i - \frac12 X_i + \frac12\right) \\
  X_i &= \frac{1}{1 + \sum_j n g_\sigma X_j\kappa_{ij} \left(e^{\beta \epsilon_{ij}}-1\right)}
\end{align}

\newcommand\epsilonassoc{\ensuremath{\varepsilon_\textit{AB}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{AB}}}
\newcommand\ncontact{\ensuremath{n_\textit{contact}}}

@@

In order to motivate the derivation of our functional for contact
density, we begin by deriving the well-known contact density for a
homogeneous hard-sphere fluid from the Carnahan-Starling free energy.
The contact density is most easily computed from the contact density
theorem, which states that the pressure on any hard surface is
determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact}.
\end{align}
Since we are interested in the contact density at the surface of the
hard spheres, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the Carnahan-Starling free energy on
hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{3} R^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi \sigma^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi \sigma^2} \frac{dA_{HS}}{dR} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

We use the White Bear version of the Fundamental-Measure Theory~(FMT)
functional published in reference~\cite{roth2002whitebear}.  The FMT
functional describes the excess free energy of a hard-sphere fluid.
This particular FMT reduces to the Carnahan-Starling equation of state
for homogeneous systems.
\begin{equation}
A_\textit{HS}[n] = k_B T \int \left(\Phi_1(\rr) + \Phi_2(\rr) + \Phi_3(\rr)\right) d\rr \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
  n_3 + (1-n_3)^2 \ln(1-n_3)
}{
  36\pi n_3^2\left( 1 - n_3 \right)^2
} ,
\end{align}
using the weighted densities
\begin{align}
  n_3(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| - R) d\rr' \\
  n_2(\rr) &= \int n(\rr') \delta(\left|\rr - \rr'\right| - R) d\rr'
\end{align}
\begin{align}
  \mathbf{n}_{V2} &= \mathbf{\nabla} n_3 , \quad
  \mathbf{n}_{V1} = \frac{\mathbf{n}_{V2}}{4\pi R} \\
  n_1 &= \frac{n_2}{4\pi R} , \quad
  n_0 = \frac{n_2}{4\pi R^2}
\end{align}

@@

%% \textcolor{red}{FIXME WRITE THIS SECTION!} The direct correlation
%% function is readily accessible from density-functional theory, as the
%% second functional derivative of the free energy:
%% \textcolor{red}{(Something like...)}
%% \begin{align}
%%   -c^{(2)}(\rr_1, \rr_2) &= \frac{1}{k_BT}\frac{\delta^2 F_{ex}}{\delta n(\rr_1) \delta n(\rr_2)}
%% \end{align}
%% The direct correlation function is related to the ordinary correlation
%% function by the Ornstein-Zernike relation
%% \begin{align}
%%   h(\rr_1, \rr'_2) &= g(\rr_1, \rr_2) - 1 \\
%%   h(\rr_1, \rr_2) &= c(\rr_1, \rr_2)
%%     + \int n(\rr_3)c(\rr_1, \rr_3) h(\rr_2, \rr_3)d\rr''
%% \end{align}
%% From these two equation, we could in principle extract the correlation
%% function evaluated at contact from the DFT free energy functional.
%% The Ornstein-Zernike equation is usually solved by taking a Fourier
%% transform of all the relevant quantities.  In the inhomogeneous case,
%% however, this approach isn't workable, since the Fourier transforms
%% don't speed things up like you'd hope, and we don't have a periodic
%% system in general.

@@

%% One reasonably simple question to ask is what the \emph{mean} contact
%% density of an inhomogeneous system is.  This comes down to simply
%% reproducing the derivation for a homogeneous system using the
%% FMT functional.
%% \begin{align}
%%   p_{HS} &= \frac{1}{N 4\pi \sigma^2} \frac{dA_{HS}}{dR} \\
%%   \ncontact &= \frac{1}{N k_BT 4\pi \sigma^2} \frac{dA_{HS}}{dR}
%% \end{align}
%% While this should give the true mean contact density for any given
%% system, this is not particularly useful, as we have no way of knowing
%% the contact density on any given sphere.  However, the mean contact
%% density comes out as an integral, so we can plot the integrand, and
%% treat it as a somewhat ill-defined local contact density.

@@

\section{Asymmetrically averaged correlation function}\label{sec:g-A}

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  g_\sigma^A(\rr) &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
Equation~\ref{eq:g-A-exact} is an exact expression for $g_\sigma^A$
given the exact functional $A_{HS}$.  In practice, we use the
approximate White Bear functional, and find an approximate expression
for $g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve more convolutions than are present in the free
energy itself, making this formulation relatively expensive.

@@

\section{Symmetrically averaged correlation function}\label{sec:g-S}

An alternate definition of the averaged correlation function for an
inhomogeneous system would be the correlation function averaged for
spheres \emph{touching at a given point}.  This would be a symmetric
measure, since both spheres involved are treated in the same way.  In
fact, the density of such spheres is just the density $n_0(\rr)$ from
FMT.  See Figure~\ref{fig:n0} for an illustration $n_0$.  Since $n_0$
is the fundamental measure of sphere surface contact, it makes sense
to define a corresponding correlation function $g_\sigma^S$.  This is
\emph{almost} what is done by Yu and

\fixme{Add qualitative description of the derivation needed to get
  $g_\sigma^S$.}

Complete expressions for $g_\sigma^S$ are included in
Appendix~\ref{appendix:g-S}.

@@

We performed a Monte-Carlo simulation of the hard sphere fluid to
measure the contact density as a function of position for a few simple
geometries.  For each scenario, we compute and compare several
quantities.  We compare the Monte Carlo density with the DFT
prediction.  We also plot Monte Carlo results for the two contact
densities: the contact density evaluated at contact, and the contact
density evaluated at the sphere location.  Alongside these, we plot
the contact density of Yu and
Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}, the contact density
of Gross~\cite{gross2009density}, and the ``contact density at
sphere'' from Section~\ref{sec:g-A}.  In principle, the
functional of Yu (FIXME) should match the CenConDensity, while the DFT
at sphere should match the ConDensity...

@@

\derivation{
\begin{widetext}
\section{The White Bear free energy for homogeneous systems}
